---
title: "week9_example"
author: "Emily Piersiak"
date: "10/20/2020"
output:
  html_document:
  toc: TRUE
  toc_float: TRUE
---
# Here is a demonstration on how to use RMarkdown
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Making a new code chunk

The shortcut key (on windows) is ctrl-alt-i.
On mac, the shortcut key is cmnd-opt-i.
If I press this key combination, RStudio will insert a new code chunk!

If I want, I can give the chunk a name and it will appear in the navigation pane.
```{r how to make a new chunk}
# this code chunk is for demonstration purposes only
```

## Reading a csv file

I'm going to read some data from a csv (comma-separated-values) file.

I can use read.csv() along with the path to the file to read the data into a data.frame.

Here's an example:

```{r read_birds_data}

require(here)
dat_birds = read.csv(here("data", "hab.sta.csv"))

```


## Previewing the contents of a data.frame

I can use the head() function to print the first few lines of a data.frame.
Here's an example with the birds data

```{r}
head(dat_birds)


```

## demo plot
```{r simple plotw}
hist(dat_birds$aspect)
```
## custom histogram               
```{r custom histogram 1}
hist(dat_birds$aspect, main = "Histogram of Study Site Aspect",
     breaks = 36)
```

