---
title: "ECo 602 Final Project"
author: "Emily Piersiak"
date: "11/27/2020"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
options(knitr.duplicate.label = TRUE)
```

Everything above this line is the header.


<!-- The following text won't be displayed in your document.  It tells R how to make nicer looking buttons for your tabbed content. -->

<style type="text/css" rel="stylesheet">

.btn {
    border-width: 0 0 0 0;
    font-weight: normal;
    text-transform: none;
}

.btn-default {
    color: #2ecc71;
    background-color: #ffffff;
    border-color: #ffffff;
}
</style>


Use this document as a template to build your final project.

If you've already created the file index.Rmd in your docs folder, you can paste everything below the header into that document.

If you don't yet have an index.Rmd file, save this document as index.Rmd in the docs folder of your main class folder.

Edit this template to fill in your own content.


# Introduction

Type a brief introduction here.


# Data Structure Functions {.tabset .tabset-pills}

In this course, we focused on a few particular functions to create and describe data structures. They include c(), data.frame(), matrix(), length(), nrow(), ncol(), and dim().


## `Vector`

The function `c()` *combines* or *concatenates* its arguments into a vector (a 1-dimensional data structure consisting of 1 or more elements).

- All of the elements must be of the same *type*.
  - I can't combine character and numeric types in the same call to `c()`

Here's two examples using numeric and character data types:

```{r}
# Create a vector of numbers:
num_vec  = c(1, 4, 8, 9, 13)

# Create a vector of characters:
char_vec = c("a", "fish", "data is cool")
```

I can show the contents of a vector by typing the name of the vector, or using the `print()` function.

```{r}
# Typing the name of the vector into the console prints the contents
num_vec

# The print() function accomplishes the same task:
print(char_vec)
```


## `Data Frame`

The function `data.frame()` creates something called a data frame, or a group of variables. This is useful to use when you import data into R for analysis.

```{r}
# Create a data frame with three columns:
data_frame1  = data.frame(1, 2, 3)

```

Data frames are created when the function `read.csv()` is used to bring a csv document into R

## `Matrix`

The function `matrix()` creates a matrix from the input set of values. 

```{r}
# Create a matrix of the list "list" with 3 rows and 3 columns:
list = c(1, 2, 3)
matrix(data = list, nrow = 3, ncol = 3, byrow = TRUE)
```

The 'nrow' argument defines the number of rows present, while 'ncol' defines the number of columns present. If the 'byrow' argument is TRUE, the matrix fills by rows. If it is FALSE, then the matrix fills by columns.

```{r}
# Create a matrix where the byrow argument is FALSE
matrix(data = list, nrow = 3, ncol = 3, byrow = FALSE)

# Create a matrix where the byrow argument is TRUE
matrix(data = list, nrow = 3, ncol = 3, byrow = TRUE)
```

## `Length`

The `length()` function can be used to find out the length of a vector, list, factor, or other R object with a defined method

```{r}
# Find the length of "length_list":
length_list = c(1, 2, 3, 4, 5, 6, 7, 8, 'a', 'b')
length(length_list)
```

## `Number of Rows`
The function `nrow()` can be used to find out how many rows are present in the datset.
```{r}
# Find the number of rows in R's base dataset "cars"
nrow(cars)
```

## `Number of Columns`

The function `ncol()` can be used to find out how many columns are present in the dataset.
```{r}
# Find the number of rows in R's base dataset "cars"
ncol(cars)
```

## `Dimensions`

The function `dim()` can be used to find the dimension of an object. The first part of the result represents rows and the second part represents columns.
```{r}
# Find the dimensions of R's base dataset "cars"
dim(cars)
```

# Subsetting Operators and Functions  {.tabset .tabset-pills}

When performing analyses, you don't always need every piece of data in the entire dataset. This is where subsetting becomes helpful: it allows you to selet particular parts of the dataset and analyze those specific parts.

## `Single Brackets`

Single brackets allow you to select certain values within a vector.

```{r}
# Create a vector
singlebracket_vector = c(1:10)

# Select the entire vector
singlebracket_vector[]

# Select the third value in the vector
singlebracket_vector[3]

# Select the values in positions 1 through 5 in the vector
singlebracket_vector[1:5]

# Select every value except for the 9th value
singlebracket_vector[-9]

# Select the values that are less than 5
singlebracket_vector[singlebracket_vector < 5]
```

## `Dollar Sign`

The dollar sign subsetting operator allows you to select a particular column within a data frame.
```{r}
# Subset the Petal.Length column from the iris dataframe
iris$Petal.Length
```

## `Subset`

The `subset()` function outputs a subset of a vector, matrix, or data frame. The first argument is the object to be subsetted from, and the second argument is what is to be subsetted.

```{r}
# Subset the rows containing information about sepals with length less than 5 from the iris dataset
subset(iris, Sepal.Length < 5)

# Subset the rows containing information about sepals with length less than 5  and width greater than 3 from the iris dataset
subset(iris, Sepal.Length < 5 & Sepal.Width > 3)
```

# Numerical Data Exploration  {.tabset .tabset-pills}

Throughout the semester, we learned a few different ways to explore data numerically. The include summary(), mean(), and sd().


## `Summary`

The function `summary()` is used to show a concise summary of an object such as a vector, list, matrix, or other object. The given summary output depends on the input object.

```{r}
# Show a summary of the list that is created below.
summary_list = c(1, 2, 3, 4, 5, 6, 7, 8, 'a', 'b')
summary(summary_list)

# Show a summary of the matrix that is created below.
summary_matrix = (matrix(data = list, nrow = 3, ncol = 3, byrow = TRUE))
summary(summary_matrix)

# Show a summary of the vector that is created below.
summary_vector = c(4, 7, 22, 57)
summary(summary_vector)

# Show a summary of the linear model that is created below.
summary_lm = lm(data = cars)
summary(summary_lm)
```


## `Mean`

The function `mean()` is used to find the arithmetic mean of a particular object.

```{r}
# Find the mean of the following vector:
mean_vector = c(4, 7, 22, 57)
mean(mean_vector)

```

## `Standard Deviation`

The function `sd()` is used to find the standard deviation of a particular object.

```{r}
# Find the standard deviation of the following vector:
sd_vector = c(4, 7, 22, 57)
sd(sd_vector)
```

# Graphical Data Exploration {.tabset .tabset-pills}

We also learned a variety of methods for graphical data exploration. The `par()` function can be used to display a number of graphs on the same panel. The funtions `plot()`, `hist()`, and `boxplot()` all create different graphs of the data.

## `Parameters`

The `par()` function allows you to set the graphical parameters. Using the `mfrow=` argument, you can create an image that shows a number of graphs side by side.

```{r}
# Create a display of two graphs side by side
par(mfrow = c(1, 2))
plot(iris$Sepal.Length, iris$Sepal.Width)
plot(iris$Sepal.Length, iris$Petal.Length)

```

## `Scatterplot`

Use the `plot()` function to create a scatterplot of two numerical datasets. 

```{r}
# Make a scatterplot of sepal length by petal length
plot(iris$Sepal.Length, iris$Petal.Length)

```

You can use a variety of arguments to modify your plot and make it easier to read, including `col=`, `pch=`, `cex=`, `main=`, `xlab=`, `ylab=`, `xlim=`, and `ylim=`.

```{r}
# Make a scatterplot of sepal length by petal length
plot(iris$Sepal.Length, iris$Petal.Length)

# Give the following:
  # a meaningful title with "main="
  # x and y axis lables with "xlab" and "ylab"
  # limits to the x and y axes with "xlim" and "ylim"
  # color with "col"
  # a nice point shape with "pch"
  # appropriately scaled text and symbols
plot(iris$Sepal.Length, iris$Petal.Length, 
     main = "Scatterplot of Iris Sepal Length by Petal Length",
     xlab = "Sepal Length (mm)",
     ylab = "Petal Length (mm)",
     xlim = c(4,8),
     ylim = c(0,8),
     col = "purple",
     pch = 6,
     cex = 1)


```



- Other color arguments are as follows:
  - col.axis (for changing the color of axis annotation) 
  - col.lab (for changing the color of the x and y labels)
  - col.main (for changing the color of the main title)
  - col.sub (for changing the color of the subtitles),
  - fg (for changing the color of the foreground - axes and boxes)
  - bg (for changing the color of the plot background).

## `Histogram`

Use the `hist()` function to create a histogram of a particular set of data. Change the number of breaks in the histogram with the `breaks` argument.

```{r}
# Create a histogram of tree height using the trees dataset with default breaks
hist(trees$Height,
     main = "Histogram of Tree Height",
     xlab = "Tree Height",
     ylab = "Frequency",
     col = "palegreen4")

# Create a histogram  of tree height with custom breaks
# Note that this isn't a great choice for break placements.
hist(trees$Height,
     main = "Histogram of Tree Height",
     xlab = "Tree Height",
     ylab = "Frequency",
     breaks = c(60,70,80,90),
     col = "palegreen4")
```

## `Boxplot`

Use `boxplot()` to create a box and whisker plot of a dataset.



```{r}
# Create a standard boxplot of survivors from the Titanic Shipwreck.
boxplot(Freq~Survived, data = Titanic,
        main = "Boxplots of Titanic Survivors",
        ylab = "Count")

# Create a conditional boxplot of survivors on sex and class from the Titanic shipwerk using the Titanic dataset
boxplot(Freq ~ Survived*Class, data = Titanic,
        main = "Boxplots of Titanic Survivors by Class",
        ylab = "Count", 
        xlab = NULL,
        names = c("Died 1st Class", "Lived 1st Class", "Died 2nd Class", "Lived 2nd Class", 
        "Died 3rd Class", "Lived 3rd Class", "Died Crew", "Lived Crew"),
        las = 2,
        cex.axis = .7
)

 # Note: the las argument changes the orientation of the axis sub-labels
```


# Distribution Functions {.tabset .tabset-pills}
There are a variety of distribution functions that we learned about in this class. There are functions for both discrete and continuous data. Both use the normal distribution, which has two parameters: mu and sigma. Changing mu, or the mean, shifts the graph to the left or right. Changing sigma, or the standard deviation, changes the width of the graph. 

## `Probability Density Function`

The `dnorm()` function is the probability density function, which is used for continuous distributions. With `dnorm()`, you can answer the question "how likely is the following value in a distribution with a particular mean and standard deviation?"

```{r}
# Find the probability of the value 1.2 showing up in a distribution with a mean of 5 and a standard deviation of 2.
dnorm(1.2, 5, 2)
```

## `Cumulative Density Function`

The `pnorm()` function is the cumulative density function, which is used with continuous distributions. With `pnorm()`, you can answer the question "how likely is it that I will choose a value of x or less from a distribution with a particular mean and standard deviation?"

```{r}
# Find the probability of picking a value of 3 or less in a distribution with a mean of 5 and a standard deviation of 2.
pnorm(3, 5, 2)
```
You can find the probability of picking a value within a particular range by subtracting one probability from another using the cumulative density function.
```{r}
# Find the probability of picking a value between 3 and 4 in a distribution with a mean of 5 and a standard deviation of 2.
pnorm(4, 5, 2) - pnorm(3, 5, 2) 
```
You can find the probability of choosing a value of x or greater by subtracting the probability of choosing that value or less from 1.
```{r}
# Find the probability of picking a value of 3 or greater in a distribution with a mean of 5 and a standard deviation of 2.
1 - pnorm(3, 5, 2) 
```
## `Quantile Function (qnorm)`

With the `qnorm()` function, also known as the quantile function, you can find the value of a particular percentile of a continuous distribution. For example, you could find the value of the 20th percentile in a distribution with a mean of x and a standard deviation of y.

```{r}
# Find the value of the 30th percentile in a distribution with a mean of 5 and a standard deviation of 2.
qnorm(.3, 5, 2)
```

## `Probability Mass Function`

- The probability mass function is used to find the probability of observing an exact value in a binomial distribution. The function `dbinom()` has the following common arguments:
  - x, or a vector of quantities
  - size, or the number of trials
  - prob, or the probability

```{r}
# Find the probability of observing a count of exactly three successes in a binomial distribution where the number of trials is 4 and the probability of success is 0.75.
dbinom(3, 4, 0.75)
 
```

## `Cumulative Mass Function`
- 
The cumulative mass function is used to find the probability of observing an exact value or less in a binomial distribution. The function `pbinom()` has the following common arguments:
  - q, or a vector of quantities
  - size, or the number of trials
  - prob, or the probability


```{r}
# Find the probability of observing a count of three successes or fewer in a binomial distribution where the number of trials is 4 and the probability of success is 0.75.
pbinom(3, 4, 0.75)
 
```

## `Quantile Function (qbinom)`

- The quantile function is used to find the value at a particular percentile in a binomial distribution. The function `qbinom()` has the following common arguments:
  - p, or the percentile
  - size, or the number of trials
  - prob, or the probability

```{r}
# Find the value of the 40th percentile in a binomial distribution with a size of 4 and a probability of success of 0.75
qbinom(.4, 4, 0.75)
 
```

# Other Functions {.tabset .tabset-pills}

These functions are incredibly important to know if you want to use R for analysis. The `library()` and `require()` functions load packages and attach any other necessary packages.

## `library()`

Add `library()` to the top of every script as a good habit to tell R that you are going to use some base functions. You can also use it to load packages for use in the analysis.

```{r}
# Load the base R library
library()

# Load the tidyverse package
library("tidyverse")
# The tidyverse is home to ggplot2 and dplyr (among many other packages), both very helpful tools for statistical analyses.
```

## `require()`
 Use `require()` to tell R that you are going to use some data that you have from a specific location, or to load a specific package.

```{r}
# Require the tidyverse package
require("tidyverse")
# Tell R that you want to use data from "here", a specified location in your project folder
require(here)
```

## `read.csv()`

Use this function to read in data and create a data frame from a .csv file. 

```{r}
# Read in the delomys dataset. First, the working directory must be set to a location where the file of interest exists.
setwd("D:/umass/eco_602_2020/ECo_602_2020/data")
read.csv('delomys.csv')
```

# Part 2: Data Anaylsis {.tabset .tabset-pills}

## `Numerical Exploration`


## `Graphical Exploration`
